---
title: "KIPRE STAFF TRAINING NEEDS ASSESSMENT (TNA) REPORT"
author: ""
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: 
  pdf_document:
    latex_engine: pdflatex
    toc: false
    toc_depth: 2

header-includes:
  - \usepackage[table]{xcolor}
  - \usepackage{multirow}
  - \usepackage{booktabs}
  - \usepackage{textgreek}
  - \usepackage{setspace}
  - \onehalfspacing
  - \usepackage{array}
  - \usepackage{colortbl}
  - \usepackage[pdfpagemode=UseNone,bookmarks=false,bookmarksopen=false,hidelinks]{hyperref}
  - \hypersetup{pdfmenubar=false,pdftoolbar=false,pdfstartview={XYZ null null 0.9}}
  - \usepackage{titling}
  - \pretitle{\begin{center}\LARGE\bfseries}
  - \posttitle{\par\vspace{1em}
    {\large Prepared by CStat. Patrick Mwaura (DSAS) \newline For \newline Head of HR, Mrs. Stella Nyambariga }\par\end{center}\vskip 7em}
  - \preauthor{}\postauthor{}
  - \predate{\begin{center}\large}\postdate{\par\end{center}}

---


```{r setup, include=FALSE}
rm(list = ls())
library(knitr)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 300
)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyverse)
library(xtable)
library(tidyverse)
library(janitor)
library(xtable)

data <- read.csv("KIPRE_TNA_25-26_Final.csv")

names(data) <- gsub("^section_[a-z]-", "", names(data)) 
names(data) <- gsub("-", ".", names(data))              
data <- as_tibble(data, .name_repair = "unique")


```

\definecolor{softred}{RGB}{255,200,200}
\definecolor{softyellow}{RGB}{255,255,200}
\definecolor{softgreen}{RGB}{200,255,200}
\definecolor{factorgray}{RGB}{220,220,220} 

\section*{Executive Summary}


\newpage


## Background

The primary purpose of this Training Needs Analysis (TNA) is to systematically assess institutional competency gaps and evaluate workforce preparedness. This evidence-based approach is designed to inform training priorities and guide strategic human capital development across the KIPRE.
The assessment was conducted using a multi-dimensional questionnaire structured into specific modules to capture a holistic view of the workforce. These sections included:

* Basic Information (Directorate, Age, and Sex of the employee)
* Job Role and Responsibilities
* Self Performances Assessing Questions
* Skills and Copetency Assessment
* Training History
* Identification of Training Needs
* Training Delivery Preferences
* Organizational Support
* Future Skills and Emerging Needs
* Additional Comments

By triangulating data from these domains, the analysis identifies not only current skill deficiencies but also the organizational support structures and delivery formats required to bridge them effectively.



## Methodology

The assessment was digitized via ODK Central and deployed from $20^{th}-28^{th}$ April of 2026. To ensure we captured a representative voice, we used a multi-channel approach—utilizing official email and professional WhatsApp groups. Participation was voluntary, positioned as a collaborative effort to co-design the organization’s future training roadmap.
Data Processing & Analysis
Once the collection window closed, raw data was ingested into R (v4.5.2). The analytical pipeline focused on cleaning, recoding qualitative responses for statistical depth, and aggregating the results to provide a granular view of needs across all Directorates and Divisions.



## Workforce profile  


A total of 80 employees, representing 42% of the 189 staff members currently at KIPRE, participated in this assessment. This respondent base spans 23 distinct divisions across seven primary Directorates, ensuring the training roadmap reflects both the core scientific mandate and the essential administrative backbone of the institute. The distribution reveals a workforce concentrated within key technical hubs, most notably the Animal Sciences Welfare and Ethics and Research and Product Development Directorates.
In the research domain, the Kenya Snakebite Research and Intervention Division and the Welfare and Ethics Division emerged as the most represented units. This engagement from frontline technical staff—supported by a demographic where 51.2% of respondents are aged between 25 and 44—provides a robust foundation for identifying gaps in laboratory biosafety and modern research protocols. While the gender distribution is relatively balanced (45% female, 55% male), distinct generational trends are evident: the female cadre is most concentrated in the 35–44 age bracket (36.1%), whereas the male workforce peaks in the 45–54 bracket (34.1%).
Administrative and strategic functions are equally well-mapped, with consistent representation from Corporate Services and the Office of the Director General. Notably, only 10% of respondents are aged 55 or older, indicating a predominantly mid-career workforce with significant long-term growth potential. This demographic profile highlights a strategic opportunity to implement cross-cutting training programmes that standardise administrative excellence and ensure the institution is prepared for future succession and evolving digital demands.


```{r workforce-table, results='asis', echo=FALSE, warning=FALSE,message=FALSE}

# 1. Prepare and aggregate data
raw_counts <- data %>%
  count(section_a.directorate, section_a.division_section) %>%
  mutate(
    Directorate = str_to_title(
      str_replace_all(section_a.directorate, "_", " ")
    ),
    
    Division = str_to_title(
      str_replace_all(section_a.division_section, "_", " ")
    )
  ) %>%
  
  arrange(Directorate) %>%
  
  select(Directorate, Division, n)

# 2. Identify where each new directorate starts
group_positions <- which(!duplicated(raw_counts$Directorate))

group_names <- raw_counts$Directorate[group_positions]

# 3. Create LaTeX row-group headers
header_commands <- paste0(
  "\\midrule\n",
  "\\multicolumn{2}{l}{\\textbf{",
  group_names,
  "}} \\\\\n",
  "\\midrule\n"
)

add_list <- list(
  pos = as.list(group_positions - 1),
  command = header_commands
)

# 4. Final display table
display_df <- raw_counts %>%
  select(Division, n) %>%
  
  rename(
    `Division / Section` = Division,
    `Staff Count` = n
  )

# 5. Print LaTeX table
print(
  xtable(
    display_df,
    caption = "Workforce distribution by directorate and division",
    align = "llc"
  ),
  
  type = "latex",
  include.rownames = FALSE,
  add.to.row = add_list,
  comment = FALSE,
  booktabs = TRUE,
  caption.placement = "top",
  
  sanitize.text.function = identity
)
```


```{r gender-age-table, results='asis', echo=FALSE, warning=FALSE, message=FALSE}


# Define desired column order
age_order <- c(
  "under_25_years",
  "25_34_years",
  "35_44_years",
  "45_54_years",
  "55+_years"
)

# Generate table
demographic_summary <- data %>%
  
  mutate(
    section_a.age_bracket = factor(
      section_a.age_bracket,
      levels = age_order,
      ordered = TRUE
    )
  ) %>%
  
  tabyl(section_a.gender, section_a.age_bracket) %>%
  
  adorn_totals(c("row", "col")) %>%
  
  adorn_percentages("row") %>%
  
  adorn_pct_formatting(digits = 1) %>%
  
  adorn_ns()

# Clean column names
colnames(demographic_summary) <- c(
  "Gender",
  "Under 25 years",
  "25-34 years",
  "35-44 years",
  "45-54 years",
  "55+ years",
  "Total"
)

# Print using xtable
print(
  xtable(
    demographic_summary,
    caption = "Gender distribution across age brackets",
    
    align = paste0(
      "l",
      paste(rep("c", ncol(demographic_summary)), collapse = "")
    )
  ),
  
  type = "latex",
  include.rownames = FALSE,
  booktabs = TRUE,
  caption.placement = "top",
  comment = FALSE
)
```


```{r grade-service-table, results='asis', echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(xtable)
library(janitor)

# define correct ordering explicitly
service_levels <- c(
  "Less Than 1 Year",
  "1 3 Years",
  "4 7 Years",
  "8 15 Years",
  "16+ Years"
)

grade_summary <- data %>%
  
  mutate(
    Job_Grade = str_to_upper(section_a.job_grade),
    Job_Grade_num = as.numeric(str_extract(Job_Grade, "\\d+")),
    
    Length_Service = section_a.length_service %>%
      str_replace_all("_", " ") %>%
      str_to_title()
  ) %>%
  
  count(Job_Grade, Job_Grade_num, Length_Service) %>%
  
  # enforce row ordering
  arrange(Job_Grade_num, Length_Service) %>%
  
  select(-Job_Grade_num) %>%
  
  pivot_wider(
    names_from = Length_Service,
    values_from = n,
    values_fill = 0
  ) %>%
  
  mutate(
    Total = rowSums(across(where(is.numeric)))
  )

# ---- FORCE COLUMN ORDER (this is the missing fix) ----
grade_summary <- grade_summary %>%
  select(
    Job_Grade,
    all_of(service_levels),
    Total
  )

# ---- FORCE ROW ORDER ----
grade_summary <- grade_summary %>%
  arrange(as.numeric(str_extract(Job_Grade, "\\d+")))

# convert to matrix so xtable cannot reorder anything
grade_matrix <- as.matrix(grade_summary)

# print
print(
  xtable(
    grade_matrix,
    caption = "Job grade distribution by length of service",
    align = c("l", rep("c", ncol(grade_matrix)))
  ),
  
  type = "latex",
  include.rownames = FALSE,
  booktabs = TRUE,
  caption.placement = "top",
  comment = FALSE
)
```

```{r, echo=FALSE,include=F, warning=FALSE, message=FALSE}


# 1. Summary by Directorate and Division
org_summary <- data %>%
  count(section_a.directorate, section_a.division_section, name = "Staff_Count") %>%
  mutate(
    Directorate = str_to_title(str_replace_all(section_a.directorate, "_", " ")),
    Division = str_to_title(str_replace_all(section_a.division_section, "_", " "))
  ) %>%
  select(Directorate, Division, Staff_Count)

# 2. Gender and Age Distribution
demographic_summary <- data %>%
  tabyl(section_a.gender, section_a.age_bracket) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns() # This shows both Count and Percentage (%)

# 3. Job Grade and Service Length
grade_summary <- data %>%
  count(section_a.job_grade, section_a.length_service) %>%
  pivot_wider(names_from = section_a.length_service, values_from = n, values_fill = 0)

# Print results to console
print("--- Staff Count by Division ---")
print(org_summary)

print("--- Gender by Age Bracket (%) ---")
print(demographic_summary)

print("--- Job Grade by Length of Service ---")
print(grade_summary)

```



### 1. Which critical competency gaps exist across directorates, divisions, and job grades, and which areas require immediate training intervention?

This question helps management identify

- High-priority skill deficiencies
- Departments with the greatest capacity gaps
- Staff categories requiring urgent support

Likely outputs

- Competency gap heatmaps
- Directorate ranking tables
- Priority training matrix
- Skill deficiency frequencies

Likely recommendations

- Targeted capacity-building programmes
- Directorate-specific training plans
- Prioritised budget allocation

```{r}
# Select, clean, and reshape data for the gap analysis
gap_analysis_data <- data %>%
  select(
    directorate = section_a.directorate,
    division = section_a.division_section,
    job_grade = section_a.job_grade,
    tech = section_d.tech_skill,
    ict = section_d.ict_skill,
    comm = section_d.comm_skill,
    leadership = section_d.leadership_management,
    policy = section_d.policy_analysis_and_implementation,
    finance = section_d.finance_management,
    customer = section_d.customer_service,
    m_and_e = section_d.monitoring_evaluation
  ) 


competencies <- c(
  "tech", "ict", "comm", "leadership",
  "policy", "finance", "customer", "m_and_e"
)
```


```{r}
gap_long <- gap_analysis_data %>%
  
  pivot_longer(
    cols = all_of(competencies),
    names_to = "competency",
    values_to = "level"
  ) %>%
  
  mutate(
    level = str_to_lower(level),
    
    gap_score = case_when(
      level == "high"   ~ 0,
      level == "medium" ~ 1,
      level == "low"    ~ 2,
      TRUE ~ NA_real_
    )
  )
```


```{r}
overall_gap_summary <- gap_long %>%
  
  group_by(competency, level) %>%
  
  summarise(
    n = n(),
    .groups = "drop"
  ) %>%
  
  group_by(competency) %>%
  
  mutate(
    percent = round(100 * n / sum(n), 1)
  ) %>%
  
  arrange(desc(percent))



```


```{r}
directorate_gap <- gap_long %>%
  
  group_by(directorate, competency) %>%
  
  summarise(
    mean_gap = mean(gap_score, na.rm = TRUE),
    low_pct  = mean(level == "low") * 100,
    med_pct  = mean(level == "medium") * 100,
    .groups = "drop"
  ) %>%
  
  arrange(desc(mean_gap))


directorate_gap %>%
  
  ggplot(aes(
    x = competency,
    y = directorate,
    fill = mean_gap
  )) +
  
  geom_tile(color = "white") +
  
  geom_text(
    aes(label = round(mean_gap, 2)),
    size = 3.5,
    color = "black"
  ) +
  
  labs(
    title = "Competency gap heatmap by directorate",
    x = "Competency",
    y = "Directorate",
    fill = "Gap score"
  ) +
  
  theme_minimal() +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )+scale_fill_gradient2(
  low = "lightblue",
  mid = "gold",
  high = "red",
  midpoint = 1
)
```


```{r}
division_gap <- gap_long %>%
  
  group_by(division, competency) %>%
  
  summarise(
    mean_gap = mean(gap_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  arrange(desc(mean_gap))

```

```{r}
job_grade_gap <- gap_long %>%
  
  group_by(job_grade, competency) %>%
  
  summarise(
    mean_gap = mean(gap_score, na.rm = TRUE),
    low_pct  = mean(level == "low") * 100,
    .groups = "drop"
  ) %>%
  
  arrange(desc(mean_gap))
```


## 2. To what extent are employees adequately prepared to meet emerging organisational, technological, and policy-related demands?

- Future workforce readiness
- Digital transformation preparedness
- Emerging competency requirements
- Institutional adaptability

Likely outputs

- Emerging skills rankings
- Workforce readiness summaries
- Future skills dashboard
- Trend analysis

Likely recommendations

- Future-focused training programmes
- Digital skills initiatives
- Leadership and innovation training
- Strategic workforce development planning

##  Key recommendations 

Priority actions   


## Conclusion   

Strategic implications 


## 7. Examplary visualisations

