---
title: "KIPRE STAFF TRAINING NEEDS ASSESSMENT (TNA) REPORT"
author: ""
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: 
  pdf_document:
    latex_engine: pdflatex
    toc: false
    toc_depth: 2

header-includes:
  - \usepackage[table]{xcolor}
  - \usepackage{multirow}
  - \usepackage{booktabs}
  - \usepackage{textgreek}
  - \usepackage{setspace}
  - \onehalfspacing
  - \usepackage{array}
  - \usepackage{colortbl}
  - \usepackage[pdfpagemode=UseNone,bookmarks=false,bookmarksopen=false,hidelinks]{hyperref}
  - \hypersetup{pdfmenubar=false,pdftoolbar=false,pdfstartview={XYZ null null 0.9}}
  - \usepackage{titling}
  - \pretitle{\begin{center}\LARGE\bfseries}
  - \posttitle{\par\vspace{1em}
    {\large Prepared by CStat. Patrick Mwaura (DSAS) \newline For \newline Head of HR, Mrs. Stella Nyambariga }\par\end{center}\vskip 7em}
  - \preauthor{}\postauthor{}
  - \predate{\begin{center}\large}\postdate{\par\end{center}}

---


```{r setup, include=FALSE}
rm(list = ls())
library(knitr)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 300
)

library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyverse)
library(xtable)
library(janitor)
library(xtable)
library(scales)
library(janitor)


data <- read.csv("KIPRE_TNA_25-26_Final.csv")

names(data) <- gsub("^section_[a-z]-", "", names(data)) 
names(data) <- gsub("-", ".", names(data))              
data <- as_tibble(data, .name_repair = "unique")
```


```{r setup2, include=FALSE}
# Clean up
data[[8]] <- recode(
  data[[8]],
  "animal_sciences_welfare_and_ethics_directorate" =
    "ASWED",
  "capacity_building_partnership_and_grant_management_directorate" =
    "CBPGMD",
  "corporate_services_directorate" =
    "CSD",
  "corporation_secretary_legal_services_directorate" =
    "CSLSD",
  "internal_audit_and_risk_assurance_directorate" =
    "IARAD",
  "office_of_the_director_general" =
    "ODG",
  "research_and_product_development_directorate" =
    "RPDD"
)
```


```{r setup3, include=FALSE}
data <- data %>%
  mutate(
    section_a.division_section = recode(
      section_a.division_section,

      "corporate_communication_division" =
        "CCD",

      "data_science_and_analytics_section" =
        "DSAS",

      "evolutionary_ecology_conservation_and_ecosystem_health_division" =
        "EECEHD",

      "finance_and_accounts_division" =
        "FAD",

      "human_resource_and_administration_division" =
        "HRAD",

      "infectious_diseases_and_one_health_division" =
        "IDOHD",

      "information_communication_technology_division" =
        "ICTD",

      "kenya_snakebite_research_and_intervention_division" =
        "KSRID",

      "laboratory_services_section" =
        "LSS",

      "national_laboratory_animal_breeding_division" =
        "NLABD",

      "non_communicable_diseases_division" =
        "NCDD",

      "partnerships_and_grants_management_division" =
        "PGMD",

      "planning_and_strategy_division" =
        "PSD",

      "reproductive_health_and_biology_division" =
        "RHBD",

      "veterinary_clinical_and_diagnostics_services_division" =
        "VCDSD",

      "welfare_and_ethics_division" =
        "WED"
    )
  )
```

\definecolor{softred}{RGB}{255,200,200}
\definecolor{softyellow}{RGB}{255,255,200}
\definecolor{softgreen}{RGB}{200,255,200}
\definecolor{factorgray}{RGB}{220,220,220} 

\section*{Executive Summary}


\newpage


## Background

The primary purpose of this Training Needs Analysis (TNA) is to systematically assess institutional competency gaps and evaluate workforce preparedness. This evidence-based approach is designed to inform training priorities and guide strategic human capital development across the KIPRE.
The assessment was conducted using a multi-dimensional questionnaire structured into specific modules to capture a holistic view of the workforce. These sections included:

* Basic Information (Directorate, Age, and Sex of the employee)
* Job Role and Responsibilities
* Self Performances Assessing Questions
* Skills and Copetency Assessment
* Training History
* Identification of Training Needs
* Training Delivery Preferences
* Organizational Support
* Future Skills and Emerging Needs
* Additional Comments

By triangulating data from these domains, the analysis identifies not only current skill deficiencies but also the organizational support structures and delivery formats required to bridge them effectively.



## Methodology

The assessment was digitized via ODK Central and deployed from $20^{th}-28^{th}$ April of 2026. To ensure we captured a representative voice, we used a multi-channel approach—utilizing official email and professional WhatsApp groups. Participation was voluntary, positioned as a collaborative effort to co-design the organization’s future training roadmap.
Data Processing & Analysis
Once the collection window closed, raw data was ingested into R (v4.5.2). The analytical pipeline focused on cleaning, recoding qualitative responses for statistical depth, and aggregating the results to provide a granular view of needs across all Directorates and Divisions.




## Workforce profile  


A total of 80 employees, representing 42% of the 189 staff members currently at KIPRE, participated in this assessment. This respondent base spans 23 distinct divisions across seven primary Directorates, ensuring the training roadmap reflects both the core scientific mandate and the essential administrative backbone of the institute. The distribution reveals a workforce concentrated within key technical hubs, most notably the Animal Sciences Welfare and Ethics and Research and Product Development Directorates.
In the research domain, the Kenya Snakebite Research and Intervention Division and the Welfare and Ethics Division emerged as the most represented units. This engagement from frontline technical staff—supported by a demographic where 51.2% of respondents are aged between 25 and 44—provides a robust foundation for identifying gaps in laboratory biosafety and modern research protocols. While the gender distribution is relatively balanced (45% female, 55% male), distinct generational trends are evident: the female cadre is most concentrated in the 35–44 age bracket (36.1%), whereas the male workforce peaks in the 45–54 bracket (34.1%).
Administrative and strategic functions are equally well-mapped, with consistent representation from Corporate Services and the Office of the Director General. Notably, only 10% of respondents are aged 55 or older, indicating a predominantly mid-career workforce with significant long-term growth potential. This demographic profile highlights a strategic opportunity to implement cross-cutting training programmes that standardise administrative excellence and ensure the institution is prepared for future succession and evolving digital demands.


```{r workforce-table, results='asis', echo=FALSE, warning=FALSE,message=FALSE}

# 1. Prepare and aggregate data
raw_counts <- data %>%
  count(section_a.directorate, section_a.division_section) %>%
  mutate(
    Directorate = str_to_title(
      str_replace_all(section_a.directorate, "_", " ")
    ),
    
    Division = str_to_title(
      str_replace_all(section_a.division_section, "_", " ")
    )
  ) %>%
  
  arrange(Directorate) %>%
  
  select(Directorate, Division, n)

# 2. Identify where each new directorate starts
group_positions <- which(!duplicated(raw_counts$Directorate))

group_names <- raw_counts$Directorate[group_positions]

# 3. Create LaTeX row-group headers
header_commands <- paste0(
  "\\midrule\n",
  "\\multicolumn{2}{l}{\\textbf{",
  group_names,
  "}} \\\\\n",
  "\\midrule\n"
)

add_list <- list(
  pos = as.list(group_positions - 1),
  command = header_commands
)

# 4. Final display table
display_df <- raw_counts %>%
  select(Division, n) %>%
  
  rename(
    `Division / Section` = Division,
    `Staff Count` = n
  )

# 5. Print LaTeX table
print(
  xtable(
    display_df,
    caption = "Workforce distribution by directorate and division",
    align = "llc"
  ),
  
  type = "latex",
  include.rownames = FALSE,
  add.to.row = add_list,
  comment = FALSE,
  booktabs = TRUE,
  caption.placement = "top",
  
  sanitize.text.function = identity
)
```


```{r gender-age-table, results='asis', echo=FALSE, warning=FALSE, message=FALSE}


# Define desired column order
age_order <- c(
  "under_25_years",
  "25_34_years",
  "35_44_years",
  "45_54_years",
  "55+_years"
)

# Generate table
demographic_summary <- data %>%
  
  mutate(
    section_a.age_bracket = factor(
      section_a.age_bracket,
      levels = age_order,
      ordered = TRUE
    )
  ) %>%
  
  tabyl(section_a.gender, section_a.age_bracket) %>%
  
  adorn_totals(c("row", "col")) %>%
  
  adorn_percentages("row") %>%
  
  adorn_pct_formatting(digits = 1) %>%
  
  adorn_ns()

# Clean column names
colnames(demographic_summary) <- c(
  "Gender",
  "Under 25 years",
  "25-34 years",
  "35-44 years",
  "45-54 years",
  "55+ years",
  "Total"
)

# Print using xtable
print(
  xtable(
    demographic_summary,
    caption = "Gender distribution across age brackets",
    
    align = paste0(
      "l",
      paste(rep("c", ncol(demographic_summary)), collapse = "")
    )
  ),
  
  type = "latex",
  include.rownames = FALSE,
  booktabs = TRUE,
  caption.placement = "top",
  comment = FALSE
)
```


```{r grade-service-table, results='asis', echo=FALSE, warning=FALSE, message=FALSE}

# define correct ordering explicitly
service_levels <- c(
  "Less Than 1 Year",
  "1 3 Years",
  "4 7 Years",
  "8 15 Years",
  "16+ Years"
)

grade_summary <- data %>%
  
  mutate(
    Job_Grade = str_to_upper(section_a.job_grade),
    Job_Grade_num = as.numeric(str_extract(Job_Grade, "\\d+")),
    
    Length_Service = section_a.length_service %>%
      str_replace_all("_", " ") %>%
      str_to_title()
  ) %>%
  
  count(Job_Grade, Job_Grade_num, Length_Service) %>%
  
  # enforce row ordering
  arrange(Job_Grade_num, Length_Service) %>%
  
  select(-Job_Grade_num) %>%
  
  pivot_wider(
    names_from = Length_Service,
    values_from = n,
    values_fill = 0
  ) %>%
  
  mutate(
    Total = rowSums(across(where(is.numeric)))
  )

# ---- FORCE COLUMN ORDER (this is the missing fix) ----
grade_summary <- grade_summary %>%
  select(
    Job_Grade,
    all_of(service_levels),
    Total
  )

# ---- FORCE ROW ORDER ----
grade_summary <- grade_summary %>%
  arrange(as.numeric(str_extract(Job_Grade, "\\d+")))

# convert to matrix so xtable cannot reorder anything
grade_matrix <- as.matrix(grade_summary)

# print
print(
  xtable(
    grade_matrix,
    caption = "Job grade distribution by length of service",
    align = c("l", rep("c", ncol(grade_matrix)))
  ),
  
  type = "latex",
  include.rownames = FALSE,
  booktabs = TRUE,
  caption.placement = "top",
  comment = FALSE
)
```

```{r, echo=FALSE,include=F, warning=FALSE, message=FALSE}


# 1. Summary by Directorate and Division
org_summary <- data %>%
  count(section_a.directorate, section_a.division_section, name = "Staff_Count") %>%
  mutate(
    Directorate = str_to_title(str_replace_all(section_a.directorate, "_", " ")),
    Division = str_to_title(str_replace_all(section_a.division_section, "_", " "))
  ) %>%
  select(Directorate, Division, Staff_Count)

# 2. Gender and Age Distribution
demographic_summary <- data %>%
  tabyl(section_a.gender, section_a.age_bracket) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns() # This shows both Count and Percentage (%)

# 3. Job Grade and Service Length
grade_summary <- data %>%
  count(section_a.job_grade, section_a.length_service) %>%
  pivot_wider(names_from = section_a.length_service, values_from = n, values_fill = 0)

# Print results to console
print("--- Staff Count by Division ---")
print(org_summary)

print("--- Gender by Age Bracket (%) ---")
print(demographic_summary)

print("--- Job Grade by Length of Service ---")
print(grade_summary)

```



### 1. Which critical competency gaps exist across directorates, divisions, and job grades, and which areas require immediate training intervention?

This question helps management identify

- High-priority skill deficiencies
- Departments with the greatest capacity gaps
- Staff categories requiring urgent support

Likely outputs

- Competency gap heatmaps
- Directorate ranking tables
- Priority training matrix
- Skill deficiency frequencies

Likely recommendations

- Targeted capacity-building programmes
- Directorate-specific training plans
- Prioritised budget allocation

```{r}
# Select, clean, and reshape data for the gap analysis
gap_analysis_data <- data %>%
  select(
    directorate = section_a.directorate,
    division = section_a.division_section,
    job_grade = section_a.job_grade,
    tech = section_d.tech_skill,
    ict = section_d.ict_skill,
    comm = section_d.comm_skill,
    leadership = section_d.leadership_management,
    policy = section_d.policy_analysis_and_implementation,
    finance = section_d.finance_management,
    customer = section_d.customer_service,
    m_and_e = section_d.monitoring_evaluation
  ) 


competencies <- c(
  "tech", "ict", "comm", "leadership",
  "policy", "finance", "customer", "m_and_e"
)
```

The baseline institutional data establishes that while the organization operates from a fundamentally stable footing, it faces a pervasive mid-level capability ceiling. Across all evaluated competencies, the workforce maintains a baseline Average Severity Score of 0.662, with a healthy 37.0% of responses indicating High Proficiency (Safe Zone) and only a minimal 3.3% falling into the Low Proficiency (Critical Alert Zone).However, a deeper look into the specific competency areas reveals that this stability is heavily anchored by general operations, specifically Comm (55.0% High), Customer (55.0% High), and baseline Tech applications (48.8% High). When these general strengths are set aside, a critical structural vulnerability emerges: a massive 59.7% company-wide majority of competencies sit clustered within the Medium Proficiency (Warning Zone). This exposure is most acute within the organization's specialized technical and regulatory pillars—specifically ICT (Information and Communications Technology) and Policy implementation—where 72.5% and 71.2% of the workforce, respectively, reside in this warning track.This profile diagnoses an organization whose staff possess excellent general baseline skills, but encounter a clear operational ceiling when handling advanced ICT core infrastructure and specialized Policy requirements. Rather than a widespread institutional crisis, the data uncovers a highly focused training mandate: strategic development resources must pivot away from generic corporate courses, and instead be aggressively targeted toward formalizing internal capacity-building pipelines for advanced ICT systems management, Monitoring & Evaluation (M&E), and Policy Implementation.

```{r, include=FALSE}
# Data Processing Pipeline
gap_long <- gap_analysis_data %>%
  pivot_longer(
    cols = all_of(competencies),
    names_to = "competency",
    values_to = "level"
  ) %>%
  mutate(
    level = str_to_lower(trimws(level)), 
    gap_score = case_when(
      level == "high"   ~ 0,
      level == "medium" ~ 1,
      level == "low"    ~ 2,
      TRUE              ~ NA_real_
    )
  )

# 1. Calculate and Clean the Metrics (Excluding Total Responses)
overall_metrics_clean <- gap_long %>%
  filter(!is.na(level), !is.na(gap_score)) %>% 
  summarise(
    `Average Severity Score` = round(mean(gap_score), 3),
    `High Proficiency (%)`   = round(mean(level == "high") * 100, 1),
    `Medium Proficiency (%)` = round(mean(level == "medium") * 100, 1),
    `Low Proficiency (%)`    = round(mean(level == "low") * 100, 1)
  )

# 2. Configure xtable for a professional PDF appearance
options(xtable.comment = FALSE) # Removes the default raw LaTeX comment banner

xtab <- xtable(
  overall_metrics_clean, 
  caption = "Overall Institutional Competency Gap Profile",
  align   = "lcccc" # Left aligns table header, centers all data columns
)

# 3. Print the LaTeX table directly into the PDF engine
print(
  xtab, 
  type = "latex", 
  include.rownames = FALSE, # Cleans out the default row index number column (1)
  booktabs = TRUE,          # Employs clean executive horizontal table lines
  table.placement = "H"     # Forces LaTeX to lock the table position exactly here
)
```
\begin{table}
\centering
\caption{Overall Institutional Competency Gap Profile} 
\begin{tabular}{cccc}
  \toprule
Average Severity Score & High Proficiency (\%) & Medium Proficiency (\%) & Low Proficiency (\%) \\ 
  \midrule
0.66 & 37.00 & 59.70 & 3.30 \\ 
   \bottomrule
\end{tabular}
\end{table}


```{r , fig.height=6, fig.width=7, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
# Color Definitions
color_high   <- "#2E7D32" # Deep Green (Safe Zone)
color_medium <- "#FBC02D" # Amber/Yellow (Warning)
color_low    <- "#C62828" # Deep Red (Severe Gap)

# 2. Corrected Data Processing Pipeline
overall_gap_summary <- gap_long %>%
  filter(!is.na(level), !is.na(gap_score)) %>% 
  group_by(competency, level) %>%
  summarise(n = n(), .groups = "drop") %>%
  
  # Calculate overall sorting weights
  group_by(competency) %>%
  mutate(
    proportion = n / sum(n),
    total_gap_weight = sum(n * case_when(level == "low" ~ 2, level == "medium" ~ 1, TRUE ~ 0)) / sum(n)
  ) %>%
  ungroup() %>%
  
  # Explicit stacking order: Low (bottom) -> Medium -> High (top)
  mutate(level = factor(level, levels = c("low", "medium", "high"))) %>%
  
  # FIXED: Arrange rows in reverse level order so cumsum() aligns with ggplot's bottom-up draw
  arrange(competency, desc(level)) %>%
  
  # Calculate coordinates
  group_by(competency) %>%
  mutate(
    label_y = cumsum(proportion) - (proportion / 2),
    competency = forcats::fct_reorder(competency, total_gap_weight, .desc = TRUE)
  ) %>%
  ungroup()

# 3. Plot Construction
ggplot(overall_gap_summary, aes(x = competency, y = proportion, fill = level)) +
  geom_col(position = "stack", width = 0.75, color = "white", lwd = 0.4) +
  
  geom_text(
    aes(
      y = label_y,
      label = scales::percent(proportion, accuracy = 0.1),
      # Ensures high visibility contrast across dark vs light shades
      color = ifelse(level == "medium", "black", "white")
    ),
    size = 3,
    fontface = "bold"
  ) +
  
  scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  
  # Clean layout text formatting for raw snake_case categories
  scale_x_discrete(
    labels = c(
      "m_and_e"    = "M & E",
      "id"         = "ID",
      "policy"     = "Policy",
      "finance"    = "Finance",
      "leadership" = "Leadership",
      "tech"       = "Tech",
      "comm"       = "Comm",
      "customer"   = "Customer"
    )
  ) +
  
  scale_fill_manual(
    values = c("low" = color_low, "medium" = color_medium, "high" = color_high),
    breaks = c("low", "medium", "high"),
    labels = c("Low (Severe Gap)", "Medium (Warning)", "High (Safe Zone)")
  ) +
  
  scale_color_identity() +
  
  labs(
    title = "Overall Competency Gap Composition Profile",
    subtitle = "Stacked from severe gap (bottom) to safe zone (top) with corrected math alignments",
    x = "Competency Core Area",
    y = "Proportional Composition Share",
    fill = "Proficiency Status"
  ) +
  
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```


```{r, fig.height=8, fig.width=7, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
# Configuration
lambda <- 5 

# Explicit, intuitive severity colors
color_safe    <- "#2E7D32" # Deep Green (High Competency = 0 Severity)
color_warning <- "#FBC02D" # Amber/Yellow (Medium Competency = 1 Severity)
color_severe  <- "#C62828" # Deep Red (Low Competency = 2 Severity)



global_mean <- mean(gap_long$gap_score, na.rm = TRUE)

directorate_gap <- gap_long %>%
  group_by(directorate, competency) %>%
  summarise(
    n = n(),
    mean_gap = mean(gap_score, na.rm = TRUE),
    
    # Shrinkage calculation
    shrink_gap = (n / (n + lambda)) * mean_gap +
                  (lambda / (n + lambda)) * global_mean,
    
    low_pct  = mean(level == "low", na.rm = TRUE) * 100,
    med_pct  = mean(level == "medium", na.rm = TRUE) * 100,
    high_pct = mean(level == "high", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    # Ordering so directorates with the WORST aggregate gaps appear at the top
    directorate = forcats::fct_reorder(directorate, shrink_gap, .fun = mean, na.rm = TRUE)
  )

# 3. Plotting Pipeline
ggplot(directorate_gap, aes(x = competency, y = directorate, fill = shrink_gap, alpha = n)) +
  geom_tile(color = "white", lwd = 0.4) +
  
  geom_text(
    aes(label = sprintf("%.2f", shrink_gap)),
    size = 3.2,
    fontface = "bold",
    color = "black",
    vjust = -0.3
  ) +
  
  geom_text(
    aes(label = paste0("H:", round(high_pct, 0), "% M:", round(med_pct, 0), "% L:", round(low_pct, 0), "%")),
    size = 1.8,
    vjust = 1.6,
    color = "grey10"
  ) +
  
  scale_alpha(range = c(0.5, 1), guide = "none") +
  
  # SEVERITY GRADIENT: Green (0) -> Yellow (1) -> Red (2)
  scale_fill_gradient2(
    low = color_safe,
    mid = color_warning,
    high = color_severe,
    midpoint = 1,
    name = "Severity Index",
    breaks = c(0, 1, 2),
    labels = c("0 (High / Safe)", "1 (Medium)", "2 (Low / Severe)")
  ) +
  
  labs(
    title = "Competency Gap Heatmap by Directorate",
    subtitle = "Shrinkage adjusted metrics (Red indicates highest deficiency severity)",
    x = "Competency Area",
    y = "Directorate",
    caption = "Values: High=0, Medium=1, Low=2. Smaller teams are safely shrunken toward global averages to eliminate sample bias."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm"),
    plot.title = element_text(face = "bold", size = 14),
    plot.caption = element_text(size = 8, hjust = 0, color = "grey40")
  )

```

```{r division_heatmap, fig.height=8, fig.width=7, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

# 1. Configuration & Colors
lambda <- 5 
color_safe    <- "#2E7D32" # Deep Green
color_warning <- "#FBC02D" # Amber/Yellow
color_severe  <- "#C62828" # Deep Red

global_mean <- mean(gap_long$gap_score, na.rm = TRUE)

# 2. Advanced Data Processing & Label Consolidation
division_gap <- gap_long %>%
  filter(!is.na(division), !is.na(competency), !is.na(level)) %>% 
  
  group_by(division, competency) %>%
  summarise(
    n = n(),
    mean_gap = mean(gap_score, na.rm = TRUE),
    shrink_gap = (n / (n + lambda)) * mean_gap +
                  (lambda / (n + lambda)) * global_mean,
    
    low_pct  = mean(level == "low", na.rm = TRUE) * 100,
    med_pct  = mean(level == "medium", na.rm = TRUE) * 100,
    high_pct = mean(level == "high", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    # Reorder Y-axis by overall deficiency weight
    division = forcats::fct_reorder(division, shrink_gap, .fun = mean, na.rm = TRUE),
    
    # FIXED: Combine both metrics into a SINGLE text label block with a clean line break (\n)
    # Format line 1: "0.77" 
    # Format line 2: "H:15 M:70 L:15"
    combined_label = paste0(
      sprintf("%.2f", shrink_gap), "\n",
      "H:", round(high_pct, 0), " M:", round(med_pct, 0), " L:", round(low_pct, 0)
    )
  )

# 3. Clean Single-Layer Plotting
ggplot(division_gap, aes(x = competency, y = division, fill = shrink_gap, alpha = n)) +
  geom_tile(color = "white", lwd = 0.5) +
  
  # FIXED: Only use ONE text geom now so overlapping is physically impossible
  geom_text(
    aes(label = combined_label),
    size = 1.8,          # Scaled for crystal-clear readability within the cells
    lineheight = 0.85,   # Slightly tightens the spacing between line 1 and line 2
    fontface = "bold",
    color = "grey15"     # High-contrast readable dark gray text
  ) +
  
  scale_alpha(range = c(0.5, 1), guide = "none") +
  
  scale_fill_gradient2(
    low = color_safe,
    mid = color_warning,
    high = color_severe,
    midpoint = 1,
    name = "Severity Index",
    breaks = c(0, 1, 2),
    labels = c("0 (High / Safe)", "1 (Medium)", "2 (Low / Severe)")
  ) +
  
  scale_x_discrete(
    labels = c(
      "m_and_e"    = "M & E",
      "id"         = "ID",
      "policy"     = "Policy",
      "finance"    = "Finance",
      "leadership" = "Leadership",
      "tech"       = "Tech",
      "comm"       = "Comm",
      "customer"   = "Customer"
    )
  ) +
  
  labs(
    title = "Competency Gap Heatmap by Division",
    subtitle = "Adjusted averages (Warm tones indicate higher training needs)",
    x = "Competency Core Area",
    y = "Division",
    caption = "Labels show: Adjusted Score (Top) / Employee Skill Breakdowns (Bottom)."
  ) +
  
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(1.2, "cm"),
    plot.title = element_text(face = "bold", size = 12),
    plot.caption = element_text(size = 7, hjust = 0, color = "grey40")
  )
```



```{r,fig.height=8, fig.width=7, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
job_grade_gap <- gap_long %>%
  
  group_by(job_grade, competency) %>%
  
  summarise(
    mean_gap = mean(gap_score, na.rm = TRUE),
    low_pct  = mean(level == "low") * 100,
    .groups = "drop"
  ) %>%
  
  arrange(desc(mean_gap))



# 1. Configuration & Shared Color Definitions
lambda <- 5 

color_safe    <- "#2E7D32" # Deep Green (High Competency = 0 Severity)
color_warning <- "#FBC02D" # Amber/Yellow (Medium Competency = 1 Severity)
color_severe  <- "#C62828" # Deep Red (Low Competency = 2 Severity)

global_mean <- mean(gap_long$gap_score, na.rm = TRUE)

# 2. Synchronized Data Processing (Aggregated by Job Grade)
job_grade_gap <- gap_long %>%
  # Filter out missing records to keep the factors tightly aligned
  filter(!is.na(job_grade), !is.na(competency), !is.na(level)) %>% 
  
  group_by(job_grade, competency) %>%
  summarise(
    n = n(),
    mean_gap = mean(gap_score, na.rm = TRUE),
    
    # Shrinkage calculation to stabilize small sample sizes at specific seniority tiers
    shrink_gap = (n / (n + lambda)) * mean_gap +
                  (lambda / (n + lambda)) * global_mean,
    
    # Distribution composition metrics per job grade cell
    low_pct  = mean(level == "low", na.rm = TRUE) * 100,
    med_pct  = mean(level == "medium", na.rm = TRUE) * 100,
    high_pct = mean(level == "high", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    # Reorder Y-axis: Seniority bands with the WORST aggregate gaps stack at the top
    job_grade = forcats::fct_reorder(job_grade, shrink_gap, .fun = mean, na.rm = TRUE),
    
    # Consolidated, non-overlapping visual text metric label
    combined_label = paste0(
      sprintf("%.2f", shrink_gap), "\n",
      "H:", round(high_pct, 0), " M:", round(med_pct, 0), " L:", round(low_pct, 0)
    )
  )

# 3. Finalized Plotting Pipeline
ggplot(job_grade_gap, aes(x = competency, y = job_grade, fill = shrink_gap, alpha = n)) +
  # Draw crisp cell partitions
  geom_tile(color = "white", lwd = 0.5) +
  
  # Combined single text layer prevents vertical overlapping bugs
  geom_text(
    aes(label = combined_label),
    size = 2.2,         
    lineheight = 0.85,  
    fontface = "bold",
    color = "grey10"
  ) +
  
  # Tile visibility mapping based on sample density (no legend)
  scale_alpha(range = c(0.5, 1), guide = "none") +
  
  # CONTINUOUS SEVERITY GRADIENT: Green (0) -> Yellow (1) -> Red (2)
  scale_fill_gradient2(
    low = color_safe,
    mid = color_warning,
    high = color_severe,
    midpoint = 1,
    name = "Severity Index",
    breaks = c(0, 1, 2),
    labels = c("0 (High / Safe)", "1 (Medium)", "2 (Low / Severe)")
  ) +
  
  # Presentation-ready name lookups for technical snake_case tokens
  scale_x_discrete(
    labels = c(
      "m_and_e"    = "M & E",
      "id"         = "ID",
      "policy"     = "Policy",
      "finance"    = "Finance",
      "leadership" = "Leadership",
      "tech"       = "Tech",
      "comm"       = "Comm",
      "customer"   = "Customer"
    )
  ) +
  
  labs(
    title = "Competency Gap Heatmap by Job Grade",
    subtitle = "Shrinkage adjusted metrics (Warm tones indicate highest deficiency severity)",
    x = "Competency Core Area",
    y = "Job Grade",
    caption = "Values: High=0, Medium=1, Low=2. Labels display: Shrunk Score / raw H-M-L percentages."
  ) +
  
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm"),
    plot.title = element_text(face = "bold", size = 14),
    plot.caption = element_text(size = 8, hjust = 0, color = "grey40")
  )

```


## 2. To what extent are employees adequately prepared to meet emerging organisational, technological, and policy-related demands?

- Future workforce readiness
- Digital transformation preparedness
- Emerging competency requirements
- Institutional adaptability

Likely outputs

- Emerging skills rankings
- Workforce readiness summaries
- Future skills dashboard
- Trend analysis

Likely recommendations

- Future-focused training programmes
- Digital skills initiatives
- Leadership and innovation training
- Strategic workforce development planning

##  Key recommendations 

Priority actions   


## Conclusion   

Strategic implications 


## 7. Examplary visualisations

