---
title: "KIPRE STAFF TRAINING NEEDS ASSESSMENT (TNA) REPORT"
author: ""
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: 
  pdf_document:
    latex_engine: pdflatex
    toc: false
    toc_depth: 2

header-includes:
  - \usepackage[table]{xcolor}
  - \usepackage{multirow}
  - \usepackage{booktabs}
  - \usepackage{textgreek}
  - \usepackage{setspace}
  - \onehalfspacing
  - \usepackage{array}
  - \usepackage{colortbl}
  - \usepackage[pdfpagemode=UseNone,bookmarks=false,bookmarksopen=false,hidelinks]{hyperref}
  - \hypersetup{pdfmenubar=false,pdftoolbar=false,pdfstartview={XYZ null null 0.9}}
  - \usepackage{titling}
  - \pretitle{\begin{center}\LARGE\bfseries}
  - \posttitle{\par\vspace{1em}
    {\large Human Resource Division \newline Prepared by CStat. Patrick Mwaura (DSAS) \newline  }\par\end{center}\vskip 7em}
  - \preauthor{}\postauthor{}
  - \predate{\begin{center}\large}\postdate{\par\end{center}}

---


```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 300
)

library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyverse)
library(xtable)
library(janitor)
library(xtable)
library(scales)
library(janitor)
library(psych)        # cronbach's alpha, PCA
library(MASS)         # polr for ordinal logistic
library(ordinal)      # optionally for clm
library(gtsummary)    # summary tables
library(flextable)    # export tables
library(DT)           # interactive tables
library(networkD3)    # Sankey diagrams
set.seed(123)         # reproducible seed


```



```{r}
# Import data
data <- read.csv("KIPRE_TNA_25-26_Final.csv")
# Clean column names: replace spaces with underscores and convert to lowercase
names(data) <- str_replace_all(names(data), " ", "_")
names(data) <- tolower(names(data))

# analysis useless variable
meta_vars <- c(
  "submissiondate", "start", "end", "today",
  "meta.instanceid", "key", "submitterid",
  "submittername", "attachmentspresent",
  "attachmentsexpected", "status",
  "reviewstate", "deviceid", "edits",
  "formversion", "notexx", "notexxx",
  "note_0", "note_1", "section_d.note_4"
)
# Remove them from the dataset
data <- data %>%
  dplyr::select(-any_of(meta_vars))
# Assign an id to each row for tracking
data <- data %>%
  mutate(id = row_number())
# Factor columns
factor_vars <- c(2:5, 8:10, 17:21,24:31, 33, 46, 47, 61,63:66)
data <- data %>%
  mutate(across(all_of(factor_vars), as.factor))

```

## Executive Summary



\newpage
## Background

The primary purpose of this Training Needs Analysis (TNA) is to systematically assess institutional competency gaps and evaluate workforce preparedness. This evidence-based approach is designed to inform training priorities and guide strategic human capital development across the KIPRE.
The assessment was conducted using a multi-dimensional questionnaire structured into specific modules to capture a holistic view of the workforce. These sections included:

* Basic Information (Directorate, Age, and Sex of the employee)
* Job Role and Responsibilities
* Self Performances Assessing Questions
* Skills and Copetency Assessment
* Training History
* Identification of Training Needs
* Training Delivery Preferences
* Organizational Support
* Future Skills and Emerging Needs
* Additional Comments

By triangulating data from these domains, the analysis identifies not only current skill deficiencies but also the organizational support structures and delivery formats required to bridge them effectively.

## Objectives

The primary goal of this assessment was to capture KIPRE employees' perspectives on their work environment, competencies, and developmental needs. By analyzing these self-reported insights, KIPRE can move away from ad-hoc training decisions and build a people-first human capital strategy. Specifically, this assessment aimed to:

* Map perceived skill gaps by uncovering where staff across different directorates, divisions, and job grades feel they lack the technical or administrative confidence to excel.
* Evaluate the felt impact of past training by documenting how effective staff believe previous workshops were and capturing the personal or structural hurdles they feel prevent them from using those skills.
* Capture bottom-up training priorities by listening directly to frontline operational needs and career progression goals to build a highly responsive institutional training roadmap.
* Tailor training delivery to staff preferences by identifying the exact learning formats, durations, and schedules that employees feel would best fit their workflows.
* Identify employee-perceived workplace bottlenecks by revealing the cultural or structural friction points that staff believe are currently limiting their productivity.
* Gauge readiness for future innovations by exploring how prepared and confident the team feels regarding emerging scientific and technological trends in research.


## Methodology

The assessment was digitized via ODK Central and deployed from $20^{th}-28^{th}$ April of 2026. To ensure we captured a representative voice, we used a multi-channel approach—utilizing official email and professional WhatsApp groups. Participation was voluntary, positioned as a collaborative effort to co-design the organization’s future training roadmap.
Data Processing & Analysis
Once the collection window closed, raw data was ingested into R (v4.5.2). The analytical pipeline focused on cleaning, recoding qualitative responses for statistical depth, and aggregating the results to provide a granular view of needs across all Directorates and Divisions.

## Results

### Workforce profile  


A total of 80 employees, representing 42% of the 189 staff members currently at KIPRE, participated in this assessment. This respondent base spans 23 distinct divisions (see Table \ref{tab:directorates}) across seven primary Directorates, ensuring the training roadmap reflects both the core scientific mandate and the essential administrative backbone of the institute. The distribution reveals a workforce concentrated within key technical hubs, most notably the Animal Sciences Welfare and Ethics and Research and Product Development Directorates.
In the research domain, the Kenya Snakebite Research and Intervention (KSRIC) Division and the Welfare and Ethics (WE) Division emerged as the most represented units. This engagement from frontline technical staff—supported by a demographic where 51.2% of respondents are aged between 25 and 44—provides a robust foundation for identifying gaps in laboratory biosafety and modern research protocols. While the gender distribution is relatively balanced (see Table \ref{tab:gender_age}) (45% female, 55% male), distinct generational trends are evident: the female cadre is most concentrated in the 35–44 age bracket (36.1%), whereas the male workforce peaks in the 45–54 bracket (34.1%).
Administrative and strategic functions are equally well-mapped, with consistent representation from Corporate Services and the Office of the Director General. Notably, only 10% of respondents are aged 55 or older, indicating a predominantly mid-career workforce with significant long-term growth potential. This demographic profile highlights a strategic opportunity to implement cross-cutting training programmes that standardise administrative excellence and ensure the institution is prepared for future succession and evolving digital demands.


```{r workforce-table, results='asis', echo=FALSE, warning=FALSE,message=FALSE}

# 1. Prepare and aggregate data
raw_counts <- data %>%
  count(section_a.directorate, section_a.division_section) %>%
  mutate(
    Directorate = str_to_title(
      str_replace_all(section_a.directorate, "_", " ")
    ),
    
    Division = str_to_title(
      str_replace_all(section_a.division_section, "_", " ")
    )
  ) %>%
  
  arrange(Directorate) %>%
  
  select(Directorate, Division, n)

# 2. Identify where each new directorate starts
group_positions <- which(!duplicated(raw_counts$Directorate))

group_names <- raw_counts$Directorate[group_positions]

# 3. Create LaTeX row-group headers
header_commands <- paste0(
  "\\midrule\n",
  "\\multicolumn{2}{l}{\\textbf{",
  group_names,
  "}} \\\\\n",
  "\\midrule\n"
)

add_list <- list(
  pos = as.list(group_positions - 1),
  command = header_commands
)

# 4. Final display table
display_df <- raw_counts %>%
  select(Division, n) %>%
  
  rename(
    `Division / Section` = Division,
    `Staff Count` = n
  )

# 5. Print LaTeX table
print(
  xtable(
    display_df,
    caption = "Workforce distribution by directorate and division", label="tab:directorates",
    align = "llc"
  ),
  
  type = "latex",
  include.rownames = FALSE,
  add.to.row = add_list,
  comment = FALSE,
  booktabs = TRUE,
  caption.placement = "top",
  
  sanitize.text.function = identity
)
```


```{r gender-age-table, results='asis', echo=FALSE, warning=FALSE, message=FALSE}


# Define desired column order
age_order <- c(
  "under_25_years",
  "25_34_years",
  "35_44_years",
  "45_54_years",
  "55+_years"
)

# Generate table
demographic_summary <- data %>%
  
  mutate(
    section_a.age_bracket = factor(
      section_a.age_bracket,
      levels = age_order,
      ordered = TRUE
    )
  ) %>%
  
  tabyl(section_a.gender, section_a.age_bracket) %>%
  
  adorn_totals(c("row", "col")) %>%
  
  adorn_percentages("row") %>%
  
  adorn_pct_formatting(digits = 1) %>%
  
  adorn_ns()

# Clean column names
colnames(demographic_summary) <- c(
  "Gender",
  "Under 25 years",
  "25-34 years",
  "35-44 years",
  "45-54 years",
  "55+ years",
  "Total"
)

# Print using xtable
print(
  xtable(
    demographic_summary,
    caption = "Gender distribution across age brackets", label="tab:gender_age",
    
    align = paste0(
      "l",
      paste(rep("c", ncol(demographic_summary)), collapse = "")
    )
  ),
  
  type = "latex",
  include.rownames = FALSE,
  booktabs = TRUE,
  caption.placement = "top",
  comment = FALSE
)
```

Based on the voluntary sample of 80 respondents, the data highlights a heavily concentrated middle-tier profile where a staggering 87.5% of all participants belong to just five core grades (NM7, NM8, NM10, NM11, and NM12) (see Table \ref{tab:lengthofservice}). This cluster is overwhelmingly driven by NM8, which represents the single largest respondent segment at 28.7% (23 individuals), followed closely by NM10 at 15.0%. Long-tenured staff are heavily represented, with over a third (36.2%) of the entire sample reporting 16 or more years of service. While this high veteran participation indicates deep institutional memory, its concentration within specific cohorts—such as NM11 where 80% of respondents have 16+ years of tenure—likely reflects a strong self-selection bias where long-serving employees felt a greater stake in completing the survey than newer hires. Furthermore, the active participation of both entry-level staff and 16+ year veterans within identical grades (such as NM8 and NM12) indicates that structural bottlenecks or career stagnation are highly visible and sensitive topics among the voluntary respondents.



```{r grade-service-table, results='asis', echo=FALSE, warning=FALSE, message=FALSE}


# define correct ordering explicitly
service_levels <- c(
  "Less Than 1 Year",
  "1 3 Years",
  "4 7 Years",
  "8 15 Years",
  "16+ Years"
)

grade_summary <- data %>%
  mutate(
    Job_Grade = str_to_upper(section_a.job_grade),
    Job_Grade_num = as.numeric(str_extract(Job_Grade, "\\d+")),
    
    Length_Service = section_a.length_service %>%
      str_replace_all("_", " ") %>%
      str_to_title()
  ) %>%
  count(Job_Grade, Job_Grade_num, Length_Service) %>%
  
  # enforce row ordering
  arrange(Job_Grade_num, Length_Service) %>%
  select(-Job_Grade_num) %>%
  
  pivot_wider(
    names_from = Length_Service,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(
    Total = rowSums(across(where(is.numeric)))
  )

# ---- FORCE COLUMN ORDER ----
grade_summary <- grade_summary %>%
  select(
    Job_Grade,
    all_of(service_levels),
    Total
  )

# ---- FORCE ROW ORDER ----
grade_summary <- grade_summary %>%
  arrange(as.numeric(str_extract(Job_Grade, "\\d+")))

# ---- ADD PERCENTAGE COLUMN (New Fix) ----
# Calculates percentage out of the cumulative sample size (sum of Total)
grand_total <- sum(grade_summary$Total)

grade_summary <- grade_summary %>%
  mutate(
    `Percentage (%)` = sprintf("%.1f", (Total / grand_total) * 100)
  )

# convert to matrix so xtable cannot reorder anything
grade_matrix <- as.matrix(grade_summary)

# print
print(
  xtable(
    grade_matrix,
    caption = "Job grade distribution by length of service (Voluntary Survey)", 
    label ="tab:lengthofservice",
    align = c("l", rep("c", ncol(grade_matrix)))
  ),
  type = "latex",
  include.rownames = FALSE,
  booktabs = TRUE,
  caption.placement = "top",
  comment = FALSE
)

```

```{r, echo=FALSE,include=F, warning=FALSE, message=FALSE}
# 1. Summary by Directorate and Division
org_summary <- data %>%
  count(section_a.directorate, section_a.division_section, name = "Staff_Count") %>%
  mutate(
    Directorate = str_to_title(str_replace_all(section_a.directorate, "_", " ")),
    Division = str_to_title(str_replace_all(section_a.division_section, "_", " "))
  ) %>%
  select(Directorate, Division, Staff_Count)

# 2. Gender and Age Distribution
demographic_summary <- data %>%
  tabyl(section_a.gender, section_a.age_bracket) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns() # This shows both Count and Percentage (%)

# 3. Job Grade and Service Length
grade_summary <- data %>%
  count(section_a.job_grade, section_a.length_service) %>%
  pivot_wider(names_from = section_a.length_service, values_from = n, values_fill = 0)

# Print results to console
print("--- Staff Count by Division ---")
print(org_summary)

print("--- Gender by Age Bracket (%) ---")
print(demographic_summary)

print("--- Job Grade by Length of Service ---")
print(grade_summary)

```

```{r setup3, include=FALSE}

data<- data %>% mutate(section_a.directorate = recode(section_a.directorate,"animal_sciences_welfare_and_ethics_directorate" =
    "ASWED",
  "capacity_building_partnership_and_grant_management_directorate" =
    "CBPGMD",
  "corporate_services_directorate" =
    "CSD",
  "corporation_secretary_legal_services_directorate" =
    "CSLSD",
  "internal_audit_and_risk_assurance_directorate" =
    "IARAD",
  "office_of_the_director_general" =
    "ODG",
  "research_and_product_development_directorate" =
    "RPDD" ) )


data <- data %>%
  mutate(
    section_a.division_section = recode(
      section_a.division_section,

      "corporate_communication_division" =
        "CCD",

      "data_science_and_analytics_section" =
        "DSAS",

      "evolutionary_ecology_conservation_and_ecosystem_health_division" =
        "EECEHD",

      "finance_and_accounts_division" =
        "FAD",

      "human_resource_and_administration_division" =
        "HRAD",

      "infectious_diseases_and_one_health_division" =
        "IDOHD",

      "information_communication_technology_division" =
        "ICTD",

      "kenya_snakebite_research_and_intervention_division" =
        "KSRID",

      "laboratory_services_section" =
        "LSS",

      "national_laboratory_animal_breeding_division" =
        "NLABD",

      "non_communicable_diseases_division" =
        "NCDD",

      "partnerships_and_grants_management_division" =
        "PGMD",

      "planning_and_strategy_division" =
        "PSD",

      "reproductive_health_and_biology_division" =
        "RHBD",

      "veterinary_clinical_and_diagnostics_services_division" =
        "VCDSD",

      "welfare_and_ethics_division" =
        "WED"
    )
  )

```


## Objective 1: Perceived skill gaps

```{r}
# Select, clean, and reshape data for the gap analysis
gap_analysis_data <- data %>%
  select(
    directorate = section_a.directorate,
    division = section_a.division_section,
    job_grade = section_a.job_grade,
    tech = section_d.tech_skill,
    ict = section_d.ict_skill,
    comm = section_d.comm_skill,
    leadership = section_d.leadership_management,
    policy = section_d.policy_analysis_and_implementation,
    finance = section_d.finance_management,
    customer = section_d.customer_service,
    m_and_e = section_d.monitoring_evaluation
  ) 


competencies <- c(
  "tech", "ict", "comm", "leadership",
  "policy", "finance", "customer", "m_and_e"
)
```

```{r overallserverity, results='asis', echo=FALSE, warning=FALSE,message=FALSE}
# Data Processing Pipeline
gap_long <- gap_analysis_data %>%
  pivot_longer(
    cols = all_of(competencies),
    names_to = "competency",
    values_to = "level"
  ) %>%
  mutate(
    level = str_to_lower(trimws(level)), 
    gap_score = case_when(
      level == "high"   ~ 0,
      level == "medium" ~ 1,
      level == "low"    ~ 2,
      TRUE              ~ NA_real_
    )
  )

# 1. Calculate and Clean the Metrics (Excluding Total Responses)
overall_metrics_clean <- gap_long %>%
  filter(!is.na(level), !is.na(gap_score)) %>% 
  summarise(
    `Average Severity Score` = round(mean(gap_score), 3),
    `High Proficiency (%)`   = round(mean(level == "high") * 100, 1),
    `Medium Proficiency (%)` = round(mean(level == "medium") * 100, 1),
    `Low Proficiency (%)`    = round(mean(level == "low") * 100, 1)
  )

# 2. Configure xtable for a professional PDF appearance
options(xtable.comment = FALSE) # Removes the default raw LaTeX comment banner

xtab <- xtable(
  overall_metrics_clean, 
  caption = "Overall Institutional Competency Gap Profile",
  label="tab:overallseverity",
  align   = "lcccc" # Left aligns table header, centers all data columns
)

# 3. Print the LaTeX table directly into the PDF engine
print(
  xtab, 
  type = "latex", 
  include.rownames = FALSE, # Cleans out the default row index number column (1)
  booktabs = TRUE,          # Employs clean executive horizontal table lines
  table.placement = "ht"     # Forces LaTeX to lock the table position exactly here
)
```


```{r , fig.height=6, fig.width=7, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
# Color Definitions
color_high   <- "#2E7D32" # Deep Green (Safe Zone)
color_medium <- "#FBC02D" # Amber/Yellow (Warning)
color_low    <- "#C62828" # Deep Red (Severe Gap)

# 2. Corrected Data Processing Pipeline
overall_gap_summary <- gap_long %>%
  filter(!is.na(level), !is.na(gap_score)) %>% 
  group_by(competency, level) %>%
  summarise(n = n(), .groups = "drop") %>%
  
  # Calculate overall sorting weights
  group_by(competency) %>%
  mutate(
    proportion = n / sum(n),
    total_gap_weight = sum(n * case_when(level == "low" ~ 2, level == "medium" ~ 1, TRUE ~ 0)) / sum(n)
  ) %>%
  ungroup() %>%
  
  # Explicit stacking order: Low (bottom) -> Medium -> High (top)
  mutate(level = factor(level, levels = c("low", "medium", "high"))) %>%
  
  # FIXED: Arrange rows in reverse level order so cumsum() aligns with ggplot's bottom-up draw
  arrange(competency, desc(level)) %>%
  
  # Calculate coordinates
  group_by(competency) %>%
  mutate(
    label_y = cumsum(proportion) - (proportion / 2),
    competency = forcats::fct_reorder(competency, total_gap_weight, .desc = TRUE)
  ) %>%
  ungroup()

# 3. Plot Construction
ggplot(overall_gap_summary, aes(x = competency, y = proportion, fill = level)) +
  geom_col(position = "stack", width = 0.75, color = "white", lwd = 0.4) +
  
  geom_text(
    aes(
      y = label_y,
      label = scales::percent(proportion, accuracy = 0.1),
      # Ensures high visibility contrast across dark vs light shades
      color = ifelse(level == "medium", "black", "white")
    ),
    size = 3,
    fontface = "bold"
  ) +
  
  scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  
  # Clean layout text formatting for raw snake_case categories
  scale_x_discrete(
    labels = c(
      "m_and_e"    = "M & E",
      "id"         = "ID",
      "policy"     = "Policy",
      "finance"    = "Finance",
      "leadership" = "Leadership",
      "tech"       = "Tech",
      "comm"       = "Comm",
      "customer"   = "Customer"
    )
  ) +
  
  scale_fill_manual(
    values = c("low" = color_low, "medium" = color_medium, "high" = color_high),
    breaks = c("low", "medium", "high"),
    labels = c("Low (Severe Gap)", "Medium (Warning)", "High (Safe Zone)")
  ) +
  
  scale_color_identity() +
  
  labs(
    title = "Overall Competency Gap Composition Profile",
    subtitle = "Stacked from severe gap (bottom) to safe zone (top) with corrected math alignments",
    x = "Competency Core Area",
    y = "Proportional Composition Share",
    fill = "Proficiency Status"
  ) +
  
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```


```{r}
# Configuration
lambda <- 5 

# Explicit, intuitive severity colors
color_safe    <- "#2E7D32" # Deep Green (High Competency = 0 Severity)
color_warning <- "#FBC02D" # Amber/Yellow (Medium Competency = 1 Severity)
color_severe  <- "#C62828" # Deep Red (Low Competency = 2 Severity)



global_mean <- mean(gap_long$gap_score, na.rm = TRUE)

directorate_gap <- gap_long %>%
  group_by(directorate, competency) %>%
  summarise(
    n = n(),
    mean_gap = mean(gap_score, na.rm = TRUE),
    
    # Shrinkage calculation
    shrink_gap = (n / (n + lambda)) * mean_gap +
                  (lambda / (n + lambda)) * global_mean,
    
    low_pct  = mean(level == "low", na.rm = TRUE) * 100,
    med_pct  = mean(level == "medium", na.rm = TRUE) * 100,
    high_pct = mean(level == "high", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    # Ordering so directorates with the WORST aggregate gaps appear at the top
    directorate = forcats::fct_reorder(directorate, shrink_gap, .fun = mean, na.rm = TRUE)
  )

# 3. Plotting Pipeline
ggplot(directorate_gap, aes(x = competency, y = directorate, fill = shrink_gap, alpha = n)) +
  geom_tile(color = "white", lwd = 0.4) +
  
  geom_text(
    aes(label = sprintf("%.2f", shrink_gap)),
    size = 3.2,
    fontface = "bold",
    color = "black",
    vjust = -0.3
  ) +
  
  geom_text(
    aes(label = paste0("H:", round(high_pct, 0), "% M:", round(med_pct, 0), "% L:", round(low_pct, 0), "%")),
    size = 1.8,
    vjust = 1.6,
    color = "grey10"
  ) +
  
  scale_alpha(range = c(0.5, 1), guide = "none") +
  
  # SEVERITY GRADIENT: Green (0) -> Yellow (1) -> Red (2)
  scale_fill_gradient2(
    low = color_safe,
    mid = color_warning,
    high = color_severe,
    midpoint = 1,
    name = "Severity Index",
    breaks = c(0, 1, 2),
    labels = c("0 (High / Safe)", "1 (Medium)", "2 (Low / Severe)")
  ) +
  
  labs(
    title = "Competency Gap Heatmap by Directorate",
    subtitle = "Shrinkage adjusted metrics (Red indicates highest deficiency severity)",
    x = "Competency Area",
    y = "Directorate",
    caption = "Values: High=0, Medium=1, Low=2. Smaller teams are safely shrunken toward global averages to eliminate sample bias."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm"),
    plot.title = element_text(face = "bold", size = 14),
    plot.caption = element_text(size = 8, hjust = 0, color = "grey40")
  )
```

```{r}
# 1. Configuration & Colors
lambda <- 5 
color_safe    <- "#2E7D32" # Deep Green
color_warning <- "#FBC02D" # Amber/Yellow
color_severe  <- "#C62828" # Deep Red

global_mean <- mean(gap_long$gap_score, na.rm = TRUE)

# 2. Advanced Data Processing & Label Consolidation
division_gap <- gap_long %>%
  filter(!is.na(division), !is.na(competency), !is.na(level)) %>% 
  
  group_by(division, competency) %>%
  summarise(
    n = n(),
    mean_gap = mean(gap_score, na.rm = TRUE),
    shrink_gap = (n / (n + lambda)) * mean_gap +
                  (lambda / (n + lambda)) * global_mean,
    
    low_pct  = mean(level == "low", na.rm = TRUE) * 100,
    med_pct  = mean(level == "medium", na.rm = TRUE) * 100,
    high_pct = mean(level == "high", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    # Reorder Y-axis by overall deficiency weight
    division = forcats::fct_reorder(division, shrink_gap, .fun = mean, na.rm = TRUE),
    
    # FIXED: Combine both metrics into a SINGLE text label block with a clean line break (\n)
    # Format line 1: "0.77" 
    # Format line 2: "H:15 M:70 L:15"
    combined_label = paste0(
      sprintf("%.2f", shrink_gap), "\n",
      "H:", round(high_pct, 0), " M:", round(med_pct, 0), " L:", round(low_pct, 0)
    )
  )

# 3. Clean Single-Layer Plotting
ggplot(division_gap, aes(x = competency, y = division, fill = shrink_gap, alpha = n)) +
  geom_tile(color = "white", lwd = 0.5) +
  
  # FIXED: Only use ONE text geom now so overlapping is physically impossible
  geom_text(
    aes(label = combined_label),
    size = 1.8,          # Scaled for crystal-clear readability within the cells
    lineheight = 0.85,   # Slightly tightens the spacing between line 1 and line 2
    fontface = "bold",
    color = "grey15"     # High-contrast readable dark gray text
  ) +
  
  scale_alpha(range = c(0.5, 1), guide = "none") +
  
  scale_fill_gradient2(
    low = color_safe,
    mid = color_warning,
    high = color_severe,
    midpoint = 1,
    name = "Severity Index",
    breaks = c(0, 1, 2),
    labels = c("0 (High / Safe)", "1 (Medium)", "2 (Low / Severe)")
  ) +
  
  scale_x_discrete(
    labels = c(
      "m_and_e"    = "M & E",
      "id"         = "ID",
      "policy"     = "Policy",
      "finance"    = "Finance",
      "leadership" = "Leadership",
      "tech"       = "Tech",
      "comm"       = "Comm",
      "customer"   = "Customer"
    )
  ) +
  
  labs(
    title = "Competency Gap Heatmap by Division",
    subtitle = "Shrinkage adjusted metrics (Warm tones indicate higher training needs)",
    x = "Competency Core Area",
    y = "Division",
    caption = "Labels show: Adjusted Score (Top) / Employee Skill Breakdowns (Bottom)."
  ) +
  
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(1.2, "cm"),
    plot.title = element_text(face = "bold", size = 12),
    plot.caption = element_text(size = 7, hjust = 0, color = "grey40")
  )

```


```{r}
job_grade_gap <- gap_long %>%
  
  group_by(job_grade, competency) %>%
  
  summarise(
    mean_gap = mean(gap_score, na.rm = TRUE),
    low_pct  = mean(level == "low") * 100,
    .groups = "drop"
  ) %>%
  
  arrange(desc(mean_gap))



# 1. Configuration & Shared Color Definitions
lambda <- 5 

color_safe    <- "#2E7D32" # Deep Green (High Competency = 0 Severity)
color_warning <- "#FBC02D" # Amber/Yellow (Medium Competency = 1 Severity)
color_severe  <- "#C62828" # Deep Red (Low Competency = 2 Severity)

global_mean <- mean(gap_long$gap_score, na.rm = TRUE)

# 2. Synchronized Data Processing (Aggregated by Job Grade)
job_grade_gap <- gap_long %>%
  # Filter out missing records to keep the factors tightly aligned
  filter(!is.na(job_grade), !is.na(competency), !is.na(level)) %>% 
  
  group_by(job_grade, competency) %>%
  summarise(
    n = n(),
    mean_gap = mean(gap_score, na.rm = TRUE),
    
    # Shrinkage calculation to stabilize small sample sizes at specific seniority tiers
    shrink_gap = (n / (n + lambda)) * mean_gap +
                  (lambda / (n + lambda)) * global_mean,
    
    # Distribution composition metrics per job grade cell
    low_pct  = mean(level == "low", na.rm = TRUE) * 100,
    med_pct  = mean(level == "medium", na.rm = TRUE) * 100,
    high_pct = mean(level == "high", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    # Reorder Y-axis: Seniority bands with the WORST aggregate gaps stack at the top
    job_grade = forcats::fct_reorder(job_grade, shrink_gap, .fun = mean, na.rm = TRUE),
    
    # Consolidated, non-overlapping visual text metric label
    combined_label = paste0(
      sprintf("%.2f", shrink_gap), "\n",
      "H:", round(high_pct, 0), " M:", round(med_pct, 0), " L:", round(low_pct, 0)
    )
  )

# 3. Finalized Plotting Pipeline
ggplot(job_grade_gap, aes(x = competency, y = job_grade, fill = shrink_gap, alpha = n)) +
  # Draw crisp cell partitions
  geom_tile(color = "white", lwd = 0.5) +
  
  # Combined single text layer prevents vertical overlapping bugs
  geom_text(
    aes(label = combined_label),
    size = 2.2,         
    lineheight = 0.85,  
    fontface = "bold",
    color = "grey10"
  ) +
  
  # Tile visibility mapping based on sample density (no legend)
  scale_alpha(range = c(0.5, 1), guide = "none") +
  
  # CONTINUOUS SEVERITY GRADIENT: Green (0) -> Yellow (1) -> Red (2)
  scale_fill_gradient2(
    low = color_safe,
    mid = color_warning,
    high = color_severe,
    midpoint = 1,
    name = "Severity Index",
    breaks = c(0, 1, 2),
    labels = c("0 (High / Safe)", "1 (Medium)", "2 (Low / Severe)")
  ) +
  
  # Presentation-ready name lookups for technical snake_case tokens
  scale_x_discrete(
    labels = c(
      "m_and_e"    = "M & E",
      "id"         = "ID",
      "policy"     = "Policy",
      "finance"    = "Finance",
      "leadership" = "Leadership",
      "tech"       = "Tech",
      "comm"       = "Comm",
      "customer"   = "Customer"
    )
  ) +
  
  labs(
    title = "Competency Gap Heatmap by Job Grade",
    subtitle = "Shrinkage adjusted metrics (Warm tones indicate highest deficiency severity)",
    x = "Competency Core Area",
    y = "Job Grade",
    caption = "Values: High=0, Medium=1, Low=2. Labels display: Shrunk Score / raw H-M-L percentages."
  ) +
  
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm"),
    plot.title = element_text(face = "bold", size = 14),
    plot.caption = element_text(size = 8, hjust = 0, color = "grey40")
  )

```


### Objevtive 2: Perceived impact of past training


### Objective 3: Perceived Training priorities

### Objective 4: Training delivery preferences


### Objective 5: Perceived workplace bottlenecks 

### Objective 6 : Innovation and innovation readiness


## Conclusion and Recommendations


